The Creation of a Programming Language

### 2022-10-10 paper.tex
add source code to paper appendix

\documentclass[12pt,a4paper,man]{apa7}
\usepackage{listings}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{dirtree}
\usepackage{times} % Use Times New Roman font
\usepackage[style=apa]{biblatex}

@@

% Defining custom commands

% include C code from file #1 from lines #2 to #3
\newcommand{\code}[3] {
    \lstset{language=C,numbers=left,basicstyle=\scriptsize,
        frame=single,caption=#1,keywordstyle=\color{blue},
        stringstyle=\color{green},commentstyle=\color{gray},
        morecomment=[l][\color{magenta}]{\#}}
    \lstinputlisting[firstnumber=#2,firstline=#2,lastline=#3]{#1}
}

@@

\subsection{The Art of Abstraction}
When dealing with problems in general, a helpful method of going about can be
to create an abstraction. This is also the case when building a programming
language interpreter. Like that, an interpreter is often times subdivided into several
components, each solving a discrete problem. One way to define
such components could be the following:
\begin{enumerate}\label{simple interpreter}
    \item Create meaningful groups of characters from the input string. This
        process is called tokenizing and is done with the component called
        tokenizer or lexer. As an example, the string 'aa+  bb' could result
        in three tokens, namely 'aa', '+' and 'bb', in case space characters
        are considered unimportant.
    \item Ordering the tokens into a so-called parse-tree or syntax-tree.
        Creating an order is important, as some calculations depend on others, e.g. multiplication
        is always done before addition. TODO:example
    \item Technically, the parse tree could now be interpreted directly, and
        an output could be produced, as the interpreter know knows what to do
        and in what order to perform computations.
\end{enumerate}
As seen above, these components, or stages as they will be called in this paper,
each input and output data, where the output of one stage is the input of the
respective next stage.

@@

\newpage\subsubsection{File Structure}
Here is a file tree overview of the project which will be referenced
several times in the following sections of this paper. In the depicted file tree
only source files are shown, no build files or executables.
\small{\setlength{\DTbaselineskip}{10pt}\dirtree{%
    .1 project root.
    .2 .gitignore.
    .2 README.md.
    .2 LICENSE.
    .2 Makefile.
    .2 mapop.gperf.
    .2 paper.tex.
    .2 bibliography.bib.
    .2 pseudocode
        .3 parser.c
    .2 src.
        .3 main.c.
        .3 interpreter.
            .4 bytecode.c.
            .4 bytecode.h.
            .4 error.
                .5 error.c.
                .5 error.h.
            .4 interpreter.c.
            .4 interpreter.h.
            .4 libraries.
                .5 libraries.c.
                .5 libraries.h.
                .5 stdlib.
                    .6 stdlib.c.
                    .6 stdlib.h.
            .4 localizer.c.
            .4 localizer.h.
            .4 mappings.
                .5 operations.c.
                .5 operations.h.
                .5 mapop.c.
                .5 mapop.h.
            .4 memory.
                .5 hashtable.c.
                .5 hashtable.h.
                .5 array.c.
                .5 array.h.
                .5 stack.c.
                .5 stack.h.
            .4 parser.c.
            .4 parser.h.
            .4 processing.
                .5 implementations.c.
                .5 implementations.h.
            .4 runtime.c.
            .4 runtime.h.
            .4 tokenizer.c.
            .4 tokenizer.h.
}}

As visible in the file tree, most of the source-code files, ending in .c,
have a corresponding header file ending in .h. This is due to the nature
of the compilation process of the C language, which is the host language
the interpreter of \name is written in.
Each header file contains only information about functions, types or variables
available to other source or header files. The actual implementations or values
are stored in the respective .c files.
TODO:explain C compilation or link example

@@

\subsubsection{Interpreter Construction}
This section will explain the process of constructing and the inner workings of
the \name interpreter. It is therefore the most complicated yet detailed section
in this paper. Examples will be made with both pseudo code and actual source
code from the project to further illustrate the approaches made. File paths
used are relative to the project's root directory.

@@

        The most important preprocessor instructions are '\#include <header.h>'
        to include various definitions from a header file and '\#define A B' to
        define A as an alias for B.
    \item \emph{Functions.} Functions in C are defined the following way:
        '<return type> <function name>([arguments]) { <function body> }'.
        In main.c there is one function. The function's name is 'main', its
        return type is 'int', it has two arguments 'argc' and 'argv' and a
        function body containing what the function does.
    \item \emph{Pointers.} Pointers are variables that store memory addresses
        of data. In C-like syntax they are defined with the star (*) symbol.
        An example from src/main.c is 'FILE *input;'. That means that input is not
        actually a FILE, but rather the memory address to where a FILE is.

        To dereference the pointer, meaning to actually retrieve the value stored
        at the memory address it contains, another star could be used as such:
        'FILE value = *input;'.
\end{itemize}

For those who are further interested in the C programming language, there are
many internet articles or tutorials going about reading and understanding or
writing and compiling C code. One such example is the article 'The C Beginner's
Handbook' provided by freeCodeCamp.org linked in the sources\footcite{freeCodeCamp}
on page \pageref{bibliography}.

@@

Following is a piece of pseudocode to better illustrate the inner workings of the
parser which is based on a Pratt-parser. 
This is only an example and not completely accurate to the actual parser used
by this project.
TODO: validate code
\code{pseudocode/parser.c}{1}{999}

@@

In src/interpreter/bytecode.c we can see, that first, the left operand and then
the right operand are added onto the stack of bytecode. Therefore, when
popping from the stack and evaluating instructions, the operands will have to
be read in reverse.
Here is an example:
\begin{enumerate}
    \item Assume the following mathematical expression in infix notation:
        \begin{center}
            \emph{'1 - 2 * 4 / (5 + 6)'}
        \end{center}
    \item In common prefix notation, also called Polish notation, the
        operators would now precede the operands as such:
        \begin{center}
            \emph{'- 1 (/ (* 2 4) (+ 5 6))'}
        \end{center}
    \item But in \name the operands are pushed onto the stack is reverse order.
        This would look like the following:
        \begin{center}
            \emph{'- (/ (+ 6 5) (* 4 2)) 1'}
        \end{center}
    \item The brackets are not needed, because for every operator, the amount
        of operands is known. This is the case as e.g. the subtraction operator
        and the negative operator, which are both a minus sign, are distinguished
        beforehand. This means the above statement would look like this:
        \begin{center}
            \emph{'- / + 6 5 * 4 2 1'}
        \end{center} 
        This is an ideal abstraction of how data 
        is represented internally in \name.
\end{enumerate}

@@

The evaluation of expressions also works in a recursive manner. Here is an 
example of \name's bytecode notation and how it would be evaluated at runtime:
\begin{itemize}
    \item Assume the input in \name's bytecode to resemble the following expression: 
        \begin{center}
            \emph{'- 6 * 3 4'}
        \end{center}
        In infix notation, this would be equivalent to 
        \begin{center}
            \emph{'(4 * 3) - 6'}.
        \end{center}
    \item The runtime will read through the bytecode left-to-right. First it will
        notice the subtraction-operator and accordingly assume that two
        operands must follow.
    \item The processing function will call itsself
        and notice the 6, a number, and therefore a value. The function simply returns the 6
        to the subtraction expression. This 6 will be the right-hand operator
        of the subtraction.
    \item The subtraction will then call the processing function again to find
        the left operand. As the processing function reads a multiplication operator,
        it now recursively reads through that expression's right and then left
        operator, which would be 3 and for respectively and compute and return
        the result of the multiplication, which is 12.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12,
        and again, calculate and return the result, which is 6.
\end{itemize}

@@

\section{Appendix}

\newpage\printbibliography[heading=bibintoc]\label{bibliography}

TODO: check correctness of all file paths!! e.g. src/...
\end{document}

>>>>

\documentclass[12pt,a4paper,man]{apa7}
\usepackage{listings}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{dirtree}
\usepackage{catchfile}
\usepackage{times} % Use Times New Roman font
\usepackage[style=apa]{biblatex}

@@

% Read contents of files.txt into \files
\CatchFileDef{\files}{files.txt}{\endlinechar=-1}

% Include C code from file #1 from lines #2 to #3
\newcommand{\code}[3] {
    \lstset{language=C,numbers=left,basicstyle=\scriptsize,
        frame=single,title=#1,keywordstyle=\color{blue},stepnumber=5,
        stringstyle=\color{green},commentstyle=\color{gray},
        morecomment=[l][\color{magenta}]{\#}}
    \lstinputlisting[firstnumber=#2,firstline=#2,lastline=#3]{#1}
}

% Center text on newline
\newcommand{\expr}[1] {
    \begin{center}
        #1
    \end{center}
}

\newcommand{\pagelabel}[1]{\phantomsection\label{#1}}

@@

\subsection{The Art of Abstraction}
When dealing with problems in general, a helpful method of going about can be
to create an abstraction. This is also the case when building a programming
language interpreter. Like that, an interpreter is often times subdivided into several
components, each solving a discrete problem. One way to define
such components could be the following:
\begin{enumerate}\pagelabel{simple interpreter}
    \item Create meaningful groups of characters from the input string. This
        process is called tokenizing and is done with the component called
        tokenizer or lexer. As an example, the string 'aa+  bb' could result
        in three tokens, namely 'aa', '+' and 'bb', in case space characters
        are considered unimportant.
    \item Ordering the tokens into a so-called parse-tree or syntax-tree.
        Creating an order is important, as some calculations depend on others, e.g. multiplication
        is always done before addition. % TODO:example
    \item Technically, the parse tree could now be interpreted directly, and
        an output could be produced, as the interpreter know knows what to do
        and in what order to perform computations.
\end{enumerate}
As seen above, these components, or stages as they will be called in this paper,
each input and output data, where the output of one stage is the input of the
respective next stage.

@@

\newpage\subsubsection{File Structure}
Here is a file tree overview of the project which will be referenced
several times in the following sections of this paper. In the depicted file tree
only source files are shown, no build files or executables.
\begin{flushleft}
    \small\setlength{\DTbaselineskip}{10pt}
    \dirtree{%
        .1 project root.
        .2 .gitignore.
        .2 README.md.
        .2 LICENSE.
        .2 Makefile.
        .2 mapop.gperf.
        .2 paper.tex.
        .2 bibliography.bib.
        .2 pseudocode.
            .3 src/interpreter/parser.c.
        .2 src.
            .3 main.c.
            .3 interpreter.
                .4 bytecode.c.
                .4 bytecode.h.
                .4 error.
                    .5 error.c.
                    .5 error.h.
                .4 interpreter.c.
                .4 interpreter.h.
                .4 libraries.
                    .5 libraries.c.
                    .5 libraries.h.
                    .5 stdlib.
                        .6 stdlib.c.
                        .6 stdlib.h.
                .4 localizer.c.
                .4 localizer.h.
                .4 mappings.
                    .5 operations.c.
                    .5 operations.h.
                    .5 mapop.c.
                    .5 mapop.h.
                .4 memory.
                    .5 hashtable.c.
                    .5 hashtable.h.
                    .5 array.c.
                    .5 array.h.
                    .5 stack.c.
                    .5 stack.h.
                .4 parser.c.
                .4 parser.h.
                .4 processing.
                    .5 implementations.c.
                    .5 implementations.h.
                .4 runtime.c.
                .4 runtime.h.
                .4 tokenizer.c.
                .4 tokenizer.h.
    }
\end{flushleft}

As visible in the file tree, most of the source-code files, ending in .c,
have a corresponding header file ending in .h. This is due to the nature
of the compilation process of the C language, which is the host language
the interpreter of \name is written in.
Each header file contains only information about functions, types or variables
available to other source or header files. The actual implementations or values
are stored in the respective .c files.
% TODO:explain C compilation or link example

@@

\subsubsection{Interpreter Construction}
This section will explain the process of constructing and the inner workings of
the \name interpreter. It is therefore the most complicated and detailed section
in this paper. Examples will be made with both pseudo code and actual source
code from the project to further illustrate the decisions made during
interpreter design. File paths
used are relative to the project's root directory.

@@

        The most common preprocessor instructions are the following.
        On one hand there is \expr{\#include <header.h>}
        to include various definitions from a header file, i.e. to paste
        its content at where the 'include' statement lies.
        On the other hand exists \expr{\#define A B} used to
        define A as an alias for B and replace every following occurrence of A
        with B.

    \item \emph{Functions.} Functions in C are defined the following way:
        \expr{<return type> <function name>([arguments]) { <function body> }}
        or with a concrete, working example:
        \expr{int myFunction(int *arg) \{ printf("arg: \%p, arg); \}.}
        In main.c there is one function. The function's name is 'main', its
        return type is 'int', it has two arguments 'argc' and 'argv' and a
        function body containing what the function does.

    \item \emph{Pointers.} Pointers are variables that store memory addresses
        of data. In C-like syntax they are defined with the star (*) symbol.
        An example from src/main.c is 'FILE *input;'. That means that input is not
        actually a FILE, but rather the memory address to where a FILE is.

        To dereference the pointer, i.e. to actually retrieve the value stored
        at the memory address it contains, another star could be used as such:
        'FILE value = *input;'.
\end{itemize}

For those who are further interested in the C programming language, there are
many internet articles or tutorials about reading and understanding or
writing and compiling C code. One such example is the article 'The C Beginner's
Handbook' provided by freeCodeCamp.org linked in the sources\footcite{freeCodeCamp}
on page \pageref{bibliography}.

@@

Following is a piece of pseudocode to better illustrate the inner workings of the
parser which is based on a Pratt-parser. 
This is only an example and not completely accurate to the actual parser used
by this project.
% TODO: validate code
\code{pseudocode/parser.c}{1}{999}

@@

In src/interpreter/bytecode.c we can see, that first, the left operand and then
the right operand are added onto the stack of bytecode. Therefore, when
popping from the stack and evaluating instructions, the operands will have to
be read in reverse.
Here is an example:
\begin{enumerate}
    \item Assume the following mathematical expression in infix notation:
        \expr{\emph{'1 - 2 * 4 / (5 + 6)'}}
    \item In common prefix notation, also called Polish notation, the
        operators would now precede the operands as such:
        \expr{\emph{'- 1 (/ (* 2 4) (+ 5 6))'}}
    \item But in \name the operands are pushed onto the stack is reverse order.
        This would look like the following:
        \expr{\emph{'- (/ (+ 6 5) (* 4 2)) 1'}}
    \item The brackets are not needed, because for every operator, the amount
        of operands is known. This is the case as e.g. the subtraction operator
        and the negative operator, which are both a minus sign, are distinguished
        beforehand. This means the above statement would look like this:
        \expr{\emph{'- / + 6 5 * 4 2 1'}}
        This is an ideal abstraction of how data 
        is represented internally in \name.
\end{enumerate}

@@

The evaluation of expressions also works in a recursive manner. Here is an 
example of \name's bytecode notation and how it would be evaluated at runtime:
\begin{itemize}
    \item Assume the input in \name's bytecode to resemble the following expression: 
        \expr{\emph{'- 6 * 3 4'}}
        In infix notation, this would be equivalent to 
        \expr{\emph{'(4 * 3) - 6'}.}
    \item The runtime will read through the bytecode left-to-right. First it will
        notice the subtraction-operator and accordingly assume that two
        operands must follow.
    \item The processing function will call itsself
        and notice the 6, a number, and therefore a value. The function simply returns the 6
        to the subtraction expression. This 6 will be the right-hand operator
        of the subtraction.
    \item The subtraction will then call the processing function again to find
        the left operand. As the processing function reads a multiplication operator,
        it now recursively reads through that expression's right and then left
        operator, which would be 3 and for respectively and compute and return
        the result of the multiplication, which is 12.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12,
        and again, calculate and return the result, which is 6.
\end{itemize}

@@

\newpage\printbibliography[heading=bibintoc]\pagelabel{bibliography}

% TODO: check correctness of all file paths!! e.g. src/...

\newpage\section{Appendix}
The appendix contains the complete source code of \name.

% Append all soruce files
\foreach \f in \files{\pagelabel{\f}\code{\f}{1}{9999}}

\end{document}
